The Impact of Urbanization, Culture, and Wealth on Public Art in Toronto

### 2024-01-25 outputs/paper/paper.qmd
Edited simulation file and made changes to final pdf.

Public art installations, or art pieces available to the general public in public spaces can be found in many urban cities across the world. Installations can be found in
New York City, Chicago, London, and even Toronto. The city of Toronto has over 400 public art installations provided by two organizations, Public Arts & Monument Collection and the Percent for Public Art Program. Public art not only enhances the urban environment, but it also brings together visitors and residents with local culture and history, creating an educational and bonding experience [@Barone].

@@

@@

In the 25-Ward model, the city of Toronto is separated into 25 neighborhoods or
voting districts. To better understand the effect of urbanization, cultural diversity, and wealth on the location of public art pieces, we compare the effect
of population, minority population, income on the number of art works across the 
25 wards. The `Ward Profiles (25-Ward Model)` data contains demographic, social, and economic information for each ward such as population, households, families, education, ethnocultural composition, spoken languages, income and housing costs. For our purpose of research, we are interested in population, minority population, and average income of each ward, so we extract these pieces of information. A sample of the cleaned dataset for the wards is shown below in @tbl-cleaned_ward. 
\vspace{10mm}

@@

### Examining Population on Public Art
As one of our variables of interest, we are determined to examine the relationship between human population and the locations of public art at the ward level. We expect population and the number of public art installations to be positively related because the prosperity and development of cities provide space and development opportunities for the development of public art [@Yin]. Public art is often a reflection of urban development, and population distribution is a direct proxy of urbanization [@Qizhi]. To visualize the relationship of interest, we plot population with the number of public art works.
```{r}
#| echo: false
#| label: pop_n
#| fig-cap: ''
#| warning: false 
  ggplot(data = art_ward_data, aes(x = population, y = n, label = WARD)) +
  geom_point() +
  geom_text_repel(hjust=-0.5, vjust=0.5) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  theme_minimal() +
  labs(x = "Population", y = "Number of Art Works") +
  scale_colour_viridis_d()

@@

### Examining Cultural Diversity on Public Art
Another explanatory variable we are interested in is the cultural diversity of a region. In this case, we visualize the relationship between minority population and the number of public art works per ward. We expect to see a positive relationship between these variables. @fig-minority-n displays the visualization. 
```{r}
#| echo: false
#| warning: false 
#| label: fig-minority-n
#| fig-cap: "Plot of Minority Population and Public Art"
  ggplot(data = art_ward_data, aes(x = minority_population, y = n, label = WARD)) +
  geom_point() +
  geom_text_repel(hjust=-0.5, vjust=0.5) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  theme_minimal() +
  labs(x = "Minority Population", y = "Number of Art Works") +
  scale_colour_viridis_d()

@@

### Examining Income and Public Art
Lastly, in this research we are hoping to look at the relationship between income by ward and the number of public art pieces. @fig-income-n plots the average household income in 2020 by ward with the number of art pieces by ward. 
```{r}
#| echo: false
#| warning: false 
#| label: fig-income-n
#| fig-cap: "Plot of Income and Public Art"
#| fig.pos: 'H'
  ggplot(data = art_ward_data, aes(x = income, y = n, label = WARD)) +
  geom_point() +
  geom_text_repel(hjust=-0.5, vjust=0.5) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  theme_minimal() +
  labs(x = "Household Income", y = "Number of Art Works") +
  scale_colour_viridis_d()

@@

# Conclusion
With the purpose of investigating the relationship between urbanization, cultural diversity, and wealth on public art in Toronto, we find a moderate positive relationship between population and number of public art installations by ward. The research finds no relationship between minority population or wealth on the number of public art pieces. suggesting that there are stronger factors that affect the locations of public art in the city. Further research could focus on the impact of population and urbanization on public art while searching for other potential variables of public art location.

>>>>

Public art installations, or art pieces available to the general public in open spaces can be found in many urban cities across the world. Installations can be found in
New York City, Chicago, London, and even Toronto. The city of Toronto has over 400 public art installations provided by two organizations, Public Arts & Monument Collection and the Percent for Public Art Program. Public art not only enhances the urban environment, but it also brings together visitors and residents with local culture and history, creating an educational and bonding experience [@Barone].

@@

Existing research on public art often focuses on the effects of public art, making public the explanatory variable. @Wright examine how public art created by Black artists in the Third Ward of Houston engages with gentrification and ghettoization. @Smith looks at the effect of public art on societal interactions that reflect complexities of Egyptian society.  Our research adds on to the literature by conducting analysis on public art as a response variable. While the societal and public art relationship could be bidirectional, we are interested in exploring the relationship between socioeconomic factors on public space usage with statistical methods.

@@

In the 25-Ward model, the city of Toronto is separated into 25 neighborhoods or
voting districts. To better understand the effect of urbanization, cultural diversity, and wealth on the location of public art pieces, we compare the effect
of population, minority population, income on the number of art works across the 
25 wards. The `Ward Profiles (25-Ward Model)` data contains demographic, social, and economic information for each ward such as population, households, families, education, ethnocultural composition, spoken languages, income and housing costs. For our purpose of research, we are interested in population, minority population, and average income of each ward.. A sample of the cleaned dataset for the wards is shown below in @tbl-cleaned_ward. 
\vspace{10mm}

@@

### Examining Population on Public Art
As one of our variables of interest, we are determined to examine the relationship between human population and the locations of public art at the ward level. We expect population and the number of public art installations to be positively related because the prosperity and development of cities provide space and development opportunities for the development of public art [@Yin]. Public art is often a reflection of urban development, and population distribution is a direct proxy of urbanization [@Qizhi]. To visualize the relationship of interest, we plot population with the number of public art works.
```{r}
#| echo: false
#| label: pop_n
#| fig-cap: "Plot of Population and Public Art by Ward"
#| warning: false 
  ggplot(data = art_ward_data, aes(x = population, y = n, label = WARD)) +
  geom_point() +
  geom_text_repel(hjust=-0.5, vjust=0.5) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  theme_minimal() +
  labs(x = "Population", y = "Number of Art Works") +
  scale_colour_viridis_d()

@@

### Examining Cultural Diversity on Public Art
Another explanatory variable we are interested in is the cultural diversity of a region. In this case, we visualize the relationship between minority population and the number of public art works per ward. We expect to see a positive relationship between these variables. @fig-minority-n displays the visualization. 
```{r}
#| echo: false
#| warning: false 
#| label: fig-minority-n
#| fig-cap: "Plot of Minority Population and Public Art by Ward"
  ggplot(data = art_ward_data, aes(x = minority_population, y = n, label = WARD)) +
  geom_point() +
  geom_text_repel(hjust=-0.5, vjust=0.5) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  theme_minimal() +
  labs(x = "Minority Population", y = "Number of Art Works") +
  scale_colour_viridis_d()

@@

### Examining Income and Public Art
Lastly, in this research we are hoping to look at the relationship between income by ward and the number of public art pieces. @fig-income-n plots the average household income in 2020 by ward with the number of art pieces by ward. 
```{r}
#| echo: false
#| warning: false 
#| label: fig-income-n
#| fig-cap: "Plot of Income and Public Art by Ward"
#| fig.pos: 'H'
  ggplot(data = art_ward_data, aes(x = income, y = n, label = WARD)) +
  geom_point() +
  geom_text_repel(hjust=-0.5, vjust=0.5) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  theme_minimal() +
  labs(x = "Household Income", y = "Number of Art Works") +
  scale_colour_viridis_d()

@@

# Conclusion
With the purpose of investigating the relationship between urbanization, cultural diversity, and wealth on public art in Toronto, we find a moderate positive relationship between population and number of public art installations by ward. The research finds no relationship between minority population or wealth on the number of public art pieces. suggesting that there may be stronger factors that affect the locations of public art in the city. Further research could focus on the impact of population and urbanization on public art while searching for other potential variables of public art location.
